---
title: "Maternal Healthcare Coverage Analysis"
author: ""
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Introduction
This report presents a population-weighted analysis of **maternal health service coverage**, using two key indicators:

- **ANC4+ (Antenatal Care Visits)**: The percentage of women aged 15–49 who received **at least four antenatal visits** during pregnancy.
- **SBA (Skilled Birth Attendance)**: The percentage of deliveries attended by **trained health personnel** (e.g., doctors, nurses, midwives).

We compare coverage levels between countries categorized as **on track** and **off track** for achieving **under-five mortality rate (U5MR)** targets. Countries with higher birth counts weigh more heavily in this analysis, ensuring that global estimates reflect where the greatest need and service impact occur.


## Data
Data Sources
We utilize three primary data sources:

1. UN World Population Estimates from 1950 to 2021 and Projections from 2022 to 2100: Provides demographic total population births.

2. UNICEF Global Dataflow: Contains maternal health indicators of ANC4 and SBA

3. U5MR Status Classification: Identifies countries on/off track for child mortality targets

## Data Cleaning Process

```{r}
# Load required packages
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(stringr)
library(rnaturalearth)
library(sf)
library(ggrepel)
#library(kableExtra)

# Function to read and clean Excel sheets
read_data <- function(file_path) {
  sheets <- c("Estimates", "Projections")
  result <- lapply(sheets, function(sheet) {
    read_excel(file_path, sheet = sheet) %>%
      filter(!is.na(...1)) %>%
      { colnames(.) <- as.character(.[1, ])
      .[-1, ]}
  })
  names(result) <- sheets
  return(result)
}

# Read and clean demographic data
data <- read_data("01_rawdata/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT_REV1.xlsx")
estimates <- data$Estimates
projections <- data$Projections

# Clean health indicator data
DATAFLOW <- read_excel("01_rawdata/GLOBAL_DATAFLOW_2018-2022.xlsx") %>% 
  filter(!is.na(`2022`)) %>% 
  select(-...3) %>% 
  rename(ANC_SBA = ...2) %>% 
  pivot_longer(cols = `2018`:`2022`, names_to = "Year", values_to = "Percentage") %>% 
  mutate(
    Percentage = ifelse(grepl("^[0-9.]+$", Percentage), Percentage, NA),
    Percentage = as.numeric(Percentage),
    ANC_SBA = case_when(
      str_detect(ANC_SBA, "Skilled birth attendant") ~ "SBA",
      str_detect(ANC_SBA, "Antenatal care") ~ "ANC4",
      TRUE ~ "OTHER")
  ) %>% 
  rename(OfficialName = `Time period`) %>% 
  pivot_wider(names_from = ANC_SBA, values_from = Percentage)

# Clean U5MR classification data
classification <- read_excel("01_rawdata/On-track and off-track countries.xlsx") %>% 
  mutate(Status.U5MR = ifelse(Status.U5MR == "Acceleration Needed", "Off-track", "On-track")) %>% 
  left_join(DATAFLOW, by = "OfficialName") %>% 
  filter(!is.na(Year)) %>% 
  mutate(across(c(ANC4, SBA), ~ as.character(map_chr(., ~ if(length(.) == 0) NA_character_ else as.character(.))))) %>% 
  mutate(across(c(ANC4, SBA), ~ na_if(., "NULL"))) %>% 
  mutate(across(c(ANC4, SBA), as.numeric)) %>% 
  filter(Year >= 2018 & Year <= 2022) %>%
  group_by(ISO3Code, OfficialName, Status.U5MR) %>%
  arrange(desc(Year)) %>%
  summarise(
    ANC4 = first(ANC4[!is.na(ANC4)]),
    SBA = first(SBA[!is.na(SBA)]),
    .groups = "drop"
  )

# Merge all datasets
unicef <- bind_rows(
  estimates %>% mutate(source = "estimates"),
  projections %>% mutate(source = "projections")) %>%
  left_join(classification %>% rename(`ISO3 Alpha-code` = ISO3Code)) %>%
  filter(
    Year %in% 2018:2022,
    !is.na(`ISO3 Alpha-code`),
    !is.na(OfficialName)
  ) %>%
  group_by(Status.U5MR) %>%
  mutate(    across( `Births (thousands)`,
      ~ replace_na(as.numeric(str_remove_all(., ",")), 0)  ) ) %>%
  ungroup() %>%
  select(`ISO3 Alpha-code`,OfficialName,Year,Status.U5MR,
         `Births (thousands)`,SBA,ANC4,`Region, subregion, country or area *`, `Parent code`) %>%
    mutate(TB_ANC4 = `Births (thousands)` * ANC4/100,
         TB_SBA = `Births (thousands)` * SBA/100,
         `Births (thousands) ANC4` = ifelse(is.na(ANC4),0,`Births (thousands)`),
         `Births (thousands) SBA` = ifelse(is.na(SBA),0,`Births (thousands)`)) 
```
We performed the following data cleaning steps:

1. **Reading Excel Sheets**: Cleaned and harmonized sheets from the World Population Prospects (WPP) file by removing header rows and reassigning column names.
2. **Transforming Indicator Data**: The raw ANC4+ and SBA indicators were reshaped for consistency. Also, we filtered only valid numeric values to clear labels.
3. **Matching Country Classifications**: Merged maternal health data with U5MR status classification based for each country, by clasifying the countries into just `On-track` and `Off-track`.
4. **Filtering Years and Selecting Most Recent Values**: Filtered to keep only coverage values from 2018 to 2022, and selecting the **most recent valid entry** of ANC4 and SBA  per country.


## Methodology


To calculate the **population-weighted coverage**, we apply the following formula:

\[ \text{Weighted Average} = \frac{\sum (\text{Coverage}_i \times \text{Births}_i)}{\sum \text{Births}_i} \]

This is applied separately for:
- **ANC4+** (4+ antenatal visits)
- **SBA** (Skilled Birth Attendance)

The goal is to estimate the global averages that account for the distribution of births, giving more weight to countries with higher fertility and birth counts.

```{r}
# Load world map data
world <- ne_countries(scale = "medium", returnclass = "sf")

# Define color palette
offtrack_color <- "#D73027"
ontrack_color <- "#1A9850"
ontrack_label_color <- "#009B95"

# Prepare mapping data
unicef_recent <- unicef %>%
  group_by(`ISO3 Alpha-code`) %>%
  filter(Year == max(Year)) %>%
  ungroup()

mapdata_filtered <- world %>%
  left_join(unicef_recent, by = c("iso_a3" = "ISO3 Alpha-code")) %>%
  mutate(
    label_color = ifelse(Status.U5MR == "On-track", "OnLabel", "OffLabel")
  )
```

## Results

The table below presents the **weighted average coverage** for each group of countries:

```{r}
percentages <- unicef %>% 
  filter(Year == 2022) %>%
  group_by(Status.U5MR) %>% 
  summarise(
    ANC4_Coverage = 100 * sum(TB_ANC4, na.rm = TRUE)/sum(`Births (thousands) ANC4`, na.rm = TRUE),
    SBA_Coverage = 100 * sum(SBA, na.rm = TRUE)/sum(`Births (thousands) SBA`, na.rm = TRUE)
  )

knitr::kable(percentages %>%
  rename(
    "U5MR Status" = Status.U5MR,
    "ANC4+ Coverage (%)" = ANC4_Coverage,
    "Skilled Birth Attendance (%)" = SBA_Coverage
  ), digits = 1)
```

- **On-track countries** have higher average coverage in both ANC4+ and SBA indicators. This suggests greater access to maternal health services where progress toward U5MR targets is being achieved.
- **Off-track countries** report lower average coverage, particularly for SBA, indicating reduced access to prenatal services and possibly systemic gaps in early maternal care.
- The coverage gap between the groups—especially pronounced in ANC4+—highlights disparities in maternal care infrastructure and service delivery.


## ANC4+ Coverage Map

This map illustrates where antenatal care coverage is high or low and how it aligns with U5MR status. Countries in grey have no available information or recent ANC4+ data.

```{r}
ggplot() +
  geom_sf(data = filter(mapdata_filtered, is.na(TB_ANC4)),
          fill = "gray80", color = "white", size = 0.2) +
  geom_sf(data = filter(mapdata_filtered, !is.na(TB_ANC4)),
          aes(fill = TB_ANC4), color = "white", size = 0.2) +
  geom_sf(data = filter(mapdata_filtered, !is.na(TB_ANC4)),
          aes(color = Status.U5MR), fill = NA, size = 0.4) +
  geom_text_repel(data = filter(mapdata_filtered, !is.na(TB_ANC4)),
                  aes(geometry = geometry, label = iso_a3, color = label_color),
                  stat = "sf_coordinates",
                  size = 3, fontface = "bold",
                  box.padding = 0.3, max.overlaps = Inf,
                  segment.color = "gray50", bg.color = "white", bg.r = 0.15,
                  seed = 42, show.legend = FALSE) +
  scale_fill_gradient(low = "#f7fcf5", high = "#00441b", 
                      name = "Total Births (thousands)", 
                      na.value = "gray80") +
  scale_color_manual(
    name = "U5MR Status",
    values = c("Off-track" = offtrack_color, "On-track" = ontrack_color,
               "OnLabel" = ontrack_label_color, "OffLabel" = offtrack_color),
    labels = c("Off-track" = "Not on track", "On-track" = "On track",
               "OnLabel" = "", "OffLabel" = ""),
    guide = guide_legend(override.aes = list(fill = NA))
  ) +
  labs(title = "Coverage of Antenatal Care (4+ Visits) Among Women Aged 15-49 in 2022",
       caption = "Source: UNICEF Data") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = "right",
        axis.text = element_blank(),
        panel.grid = element_blank())
```

- Higher ANC4+ coverage is mostly clustered in countries categorized as on-track, especially in regions like Latin America and parts of Asia.
- Many countries in Africa show low ANC4+ coverage and are predominantly off-track.
- The presence of large gaps in data availability in several low-income countries further highlights areas requiring data system strengthening.

## Skilled Birth Attendance

```{r}
ggplot() +
  geom_sf(data = filter(mapdata_filtered, is.na(TB_SBA)),
          fill = "gray80", color = "white", size = 0.2) +
  geom_sf(data = filter(mapdata_filtered, !is.na(TB_SBA)),
          aes(fill = TB_SBA), color = "white", size = 0.2) +
  geom_sf(data = filter(mapdata_filtered, !is.na(TB_SBA)),
          aes(color = Status.U5MR), fill = NA, size = 0.4) +
  geom_text_repel(data = filter(mapdata_filtered, !is.na(TB_SBA)),
                  aes(geometry = geometry, label = iso_a3, color = label_color),
                  stat = "sf_coordinates",
                  size = 3, fontface = "bold",
                  box.padding = 0.3, max.overlaps = Inf,
                  segment.color = "gray50", bg.color = "white", bg.r = 0.15,
                  seed = 42, show.legend = FALSE) +
  scale_fill_gradient(low = "#f7fcf5", high = "#00441b", 
                      name = "Total Births (thousands)", 
                      na.value = "gray80") +
  scale_color_manual(
    name = "U5MR Status",
    values = c("Off-track" = offtrack_color, "On-track" = ontrack_color,
               "OnLabel" = ontrack_label_color, "OffLabel" = offtrack_color),
    labels = c("Off-track" = "Not on track", "On-track" = "On track",
               "OnLabel" = "", "OffLabel" = ""),
    guide = guide_legend(override.aes = list(fill = NA))
  ) +
  labs(title = "Coverage of Deliveries Attended by Skilled Health Personnel in 2022",
       caption = "Source: UNICEF Data") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = "right",
        axis.text = element_blank(),
        panel.grid = element_blank())
```

- SBA coverage is generally higher and more consistent across both groups compared to ANC4+.
- Nevertheless, several off-track countries—particularly in Central Africa—remain significantly behind in ensuring deliveries are attended by skilled professionals.
- The map shows a narrower global disparity in SBA than ANC4+, but the shortfall in off-track countries is bigger.


## Skilled Birth Attendance Anual Variations

```{r}
unicefVariation = unicef %>% 
  left_join(estimates %>%
              filter(Type == "Subregion") %>%
              distinct(`Region, subregion, country or area *`, `Location code`, `Parent code`) %>%
              rename(Subregion = `Region, subregion, country or area *`) %>%
              left_join(estimates %>%
                          filter(Type == "Region") %>%
                          distinct(`Region, subregion, country or area *`, `Location code`) %>%
                          rename(Region = `Region, subregion, country or area *`,
                                 `Parent code` = `Location code`)) %>%
              select(-`Parent code`) %>%
              rename(`Parent code` = `Location code`)) %>% 
  group_by(Year, Region, Status.U5MR) %>%
  summarise(Total = sum(TB_SBA)) %>%
  group_by(Region, Status.U5MR) %>%
  mutate(Total_diff = Total - lag(Total),
         Total_pct_change = (Total - lag(Total)) / lag(Total) * 100,
         Region = ifelse(is.na(Region), "NORTHERN AMERICA", Region)) %>%
  filter(!is.na(Total_pct_change))


#------------------------
# Plot 2: Time series of % population change
#------------------------
ggplot(unicefVariation, aes(x = as.integer(Year), y = Total_pct_change, color = Region, group = Region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = paste0(round(Total_pct_change, 1), "%")),
                  size = 4,
                  fontface = "bold",
                  max.overlaps = 50,
                  show.legend = FALSE) +
  facet_wrap(~ Status.U5MR, scales = "fixed", labeller = as_labeller(c(
    "On-track" = "On-track",
    "Off-track" = "Off-track"))) +
  labs(
    title = "Anual % Change in Coverage of Deliveries Attended by Skilled Health Personnel",
    x = "Year",
    y = "% Change",
    color = "Region",
    caption = "Source: UNICEF Population Data, 2018–2022") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 13),
        strip.text = element_text(size = 14, face = "bold"),
        legend.title = element_text(face = "bold"))

```

- Northern America experienced the strongest bounce: after a –1.2% drop in 2019, it surged to +1.4% in 2021, followed by a slight dip to +0.2% in 2022.

- Africa and Oceania had positive trends, though Africa declined again to –1.2% by 2022.

- Europe and Latin America improved modestly, reaching –1.1% to –0.9% change in 2022.

- Asia (not shown explicitly) may be reflected within Oceania or other regions if grouped.

## Conclusions

This analysis reveals contrasts in maternal health service access across U5MR performance groups. Key takeaways include:

- Countries **on track to meet U5MR goals** have significantly higher population-weighted coverage for both ANC4+ and SBA indicators.
- Countries **off track** display lower coverage and represent a major gap in global maternal healthcare delivery.
- Using **population-weighted metrics** helps better reflect needs and disparities than simple averages.
- **ANC4+ coverage** is particularly lower in off-track countries, suggesting a more urgent need to improve prenatal service access compared to SBA.
- **Recent trends in SBA coverage (2019–2022)** reveal that **on-track countries** show more **resilience and recovery capacity**, with several regions improving coverage after initial declines, while **off-track countries** exhibit **flat or negative growth**, indicating deeper systemic barriers.


